---
title: 'Interactive Data Visualization with plotly in R '
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    css: doc.css
    df_print: paged
  pdf_document: default
---


```{r eval=FALSE,warning = FALSE, message = FALSE}
# used libraries
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("plotly")) install.packages("plotly")
# R package DT provides an R interface to the JavaScript library DataTables.
if (!require("DT")) install.packages("DT")
# geom_mark_hull
if (!require("concaveman")) install.packages("concaveman")
if (!require("ggforce")) install.packages("ggforce")


# more libraries
if (!require("listviewer")) install.packages("listviewer")
if (!require("lubridate")) install.packages("lubridate")
if (!require("forcats")) install.packages("forcats")
if (!require("maps")) install.packages("maps")
if (!require("mvtnorm")) install.packages("mvtnorm")
if (!require("gapminder")) install.packages("gapminder")
if (!require("hexbin")) install.packages("hexbin")
if (!require("Hmisc")) install.packages("Hmisc")
if (!require("leaflet")) install.packages("leaflet")

if (!require("plotlyBook")) install.packages("plotlyBook")

```
```{r setup, include=FALSE,warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
library(ggplot2)
library(plotly)
library(ggforce)
library(dplyr)
library(lubridate)
```

# plotly - Documentation

* Documentation:
  * [https://plotly.com/r/](https://plotly.com/r/)
* Book:
  * Interactive web-based data visualization with R, plotly, and shiny
by Carson Sievert [https://plotly-r.com/](https://plotly-r.com/)

This document is essentially a brief notes of the book above.

# Linking Multiple Views - Client Side Linking



* Several high level R libraries provides highlevel interface for JavaScript/html UI.
  * `htmltools` - Tools for creating, manipulating, and writing HTML from R.
  * `htmlwidget` (on top `htmltools`) provides a framework for easily creating 
    R bindings to JavaScript libraries. Widgets created using the framework can be:
    * Used at the R console for data analysis just like conventional R plots (via RStudio Viewer)
    * Seamlessly embedded within R Markdown documents and Shiny web applications.
    * Saved as standalone web pages for ad-hoc sharing via email, Dropbox, etc.
  * `crosstalk` - provides building blocks for allowing HTML widgets to communicate 
     with each other, with Shiny or without (i.e. static .html files).
  * `shiny` - Web Application Framework for R, Makes it incredibly easy to 
     build interactive web applications with R
  * `plotly` - interactive data visualization
  * other htmlwidgets: `DT`, `leaflet`, ...
    * `DT` - dataframe visualization
    * `leaflet` - geo maps

  
Client Side Linking means that (built-in) Browser handle all output/interactions using JavaScript/html tech.

## plotly’s framework for highlight events

* A highlight event dims the opacity of existing marks, then adds an additional graphical layer representing the selection.


Two components for highlighting:

* `highlight_key`() function - set the selection option for higlight
  * specify selection variable through formula.
* `highlight`()  this section shows you how to control how queries are triggered and visually rendered via the highlight() function."""
  * Some impotent arguments:
    * `on`/`off` - on which event to turn on/off highlighting, some possible events: "plotly_hover", "plotly_click", "plotly_doubleclick"
    * `selectize` - turn on selectize.js powered dropdown widget
    * `dynamic` - should a widget for changing selection colors be included (add `color`)
    * `persistent` - should selections persist
    
```{r}
library(plotly)
head(mtcars)

mtcars_key <- highlight_key(mtcars, ~cyl)
  
p <- plot_ly(mtcars_key,
  x = ~wt, y = ~mpg, text = ~cyl, mode = "markers+text", 
  textposition = "top", hoverinfo = "x+y"
)
highlight(p, on = "plotly_click", off = "plotly_doubleclick")
```


```{r}
library(plotly)
head(mtcars)
mtcars %>%
  highlight_key(~cyl) %>%
  plot_ly(
    x = ~wt, y = ~mpg, text = ~cyl, mode = "markers+text", 
    textposition = "top", hoverinfo = "x+y"
  ) %>%
  highlight(on = "plotly_hover", off = "plotly_doubleclick")
```
```{r}
# load the `txhousing` dataset
data(txhousing, package = "ggplot2")

highlight_key(txhousing, ~city) %>% # declare `city` as highlight key
  plot_ly(color = I("black")) %>% 
  group_by(city) %>%
  add_lines(x = ~date, y = ~median)# create a time series of median house price
```

 As it turns out, plotly makes it easy to add a selectize.js powered dropdown widget for querying by name (aka indirect manipulation) by setting selectize = TRUE.
 
 

```{r}
highlight_key(txhousing, ~city) %>% # declare `city` as highlight key
  plot_ly(color = I("black")) %>% 
  group_by(city) %>%
  add_lines(x = ~date, y = ~median) %>%# create a time series of median house price
  highlight(
    on = "plotly_click", 
    selectize = TRUE, 
    dynamic = TRUE, 
    persistent = TRUE
  )
```


```{r}

# remember, `base` is a plotly object, but we can use dplyr verbs to
# manipulate the input data 
# (`txhousing` with `city` as a grouping and querying variable)

base <- highlight_key(txhousing, ~city) %>% # declare `city` as highlight key
  plot_ly(color = I("black")) %>% 
  group_by(city)

dot_plot <- base %>%
  summarise(miss = sum(is.na(median))) %>%
  filter(miss > 0) %>%
  add_markers(
    x = ~miss, 
    y = ~forcats::fct_reorder(city, miss), 
    hoverinfo = "x+y"
  ) %>%
  layout(
    xaxis = list(title = "Number of months missing"),
    yaxis = list(title = "")
  ) 
time_series <- base %>%
  group_by(city) %>%
  add_lines(x = ~date, y = ~median) 

subplot(dot_plot, time_series, widths = c(.2, .8), titleX = TRUE) %>%
  layout(showlegend = FALSE) %>%
  highlight(on = "plotly_click", dynamic = TRUE, selectize = TRUE)
```

```{r}
hist <- add_histogram(
  base,
  x = ~median, 
  histnorm = "probability density"
)
subplot(time_series, hist, nrows = 2) %>%
  layout(barmode = "overlay", showlegend = FALSE) %>%
  highlight(
    dynamic = TRUE, 
    selectize = TRUE, 
    selected = attrs_selected(opacity = 0.3)
  )

```

Other htmlwidgets can be combined together


```{r}
library(plotly)

m <- highlight_key(mpg)
p <- ggplot(m, aes(displ, hwy)) + geom_point()
gg <- highlight(ggplotly(p), "plotly_selected")

# arrange widgets in columns (from crosstalk)
crosstalk::bscols(gg, DT::datatable(m))
```

## plotly’s framework for filter events

* A filter event completely remove existing marks and rescales axes to the remaining data.

filter_select(),

```{r}
library(crosstalk)

# highlight
tx <- highlight_key(txhousing, ~city, "Select a city")
gg <- ggplot(tx) + geom_line(aes(date, median, group = city))

highlight_p <- highlight(
  ggplotly(gg, tooltip = "city"), 
  selectize = TRUE, persistent = TRUE
)

# filter

# generally speaking, use a "unique" key for filter, 
# especially when you have multiple filters!
tx2 <- highlight_key(txhousing)
gg2 <- ggplot(tx2) + geom_line(aes(date, median, group = city))

# arrange widgets in columns (from crosstalk)
filter_p <- bscols(
  filter_select("id", "Select a city", tx2, ~city),
  ggplotly(gg2, dynamicTicks = TRUE),
  widths = c(12, 12)
)



# arrange widgets in columns (from crosstalk)
bscols(highlight_p, filter_p)
```

```{r}

library(crosstalk)

tx <- highlight_key(txhousing)

widgets <- bscols(
  widths = c(12, 12, 12),
  filter_select("city", "Cities", tx, ~city),
  filter_slider("sales", "Sales", tx, ~sales),
  filter_checkbox("year", "Years", tx, ~year, inline = TRUE)
)
bscols(
  widths = c(4, 8), widgets, 
  plot_ly(tx, x = ~date, y = ~median, showlegend = FALSE) %>% 
    add_lines(color = ~city, colors = "black")
)
```

 filter and highlight events can work in conjunction with various htmlwidgets. 
 In fact, since the semantics of filter are more well-defined than highlight, 
 linking filter events across htmlwidgets via crosstalk should generally be more well-supported.

```{r}

library(leaflet)

eqs <- highlight_key(quakes)
stations <- filter_slider(
  "station", "Number of Stations", 
  eqs, ~stations
)

p <- plot_ly(eqs, x = ~depth, y = ~mag) %>% 
  add_markers(alpha = 0.5) %>% 
  highlight("plotly_selected")

map <- leaflet(eqs) %>% 
  addTiles() %>% 
  addCircles()

bscols(
  widths = c(6, 6, 3), 
  p, map, stations
)
```

```{r}
library(gapminder)
g <- highlight_key(gapminder, ~country)
continent_filter <- filter_select(
  "filter", "Select a country", 
  g, ~continent
)

p <- plot_ly(g) %>%
  group_by(country) %>%
  add_lines(x = ~year, y = ~lifeExp, color = ~continent) %>%
  layout(xaxis = list(title = "")) %>%
  highlight(selected = attrs_selected(showlegend = FALSE))

bscols(continent_filter, p, widths = 12)
```
# Animation

```{r}
g <- highlight_key(gapminder, ~continent)
gg <- ggplot(g, aes(gdpPercap, lifeExp, 
  color = continent, frame = year)) +
  geom_point(aes(size = pop, ids = country)) +
  geom_smooth(se = FALSE, method = "lm") +
  scale_x_log10()
highlight(ggplotly(gg), "plotly_click")

```

```{r}
library(gapminder)

# gapKey <- highlight_key(gap, ~country)
gap <- gapminder %>% mutate(country=forcats::fct_reorder(as.character(country), pop)) # forcats::fct_reorder for factor reorder
levels(gap$country)
plot_ly(gap, y = ~factor(country), x = ~pop, hoverinfo = "x") %>%
  add_markers(alpha = 0.1, color = I("black"))
```

```{r}
gap <- gapminder %>% mutate(country=forcats::fct_reorder(as.character(country), pop)) 
# forcats::fct_reorder for factor reorder

gapKey <- highlight_key(gap, ~country)

p1 <- plot_ly(gap, y = ~country, x = ~pop, hoverinfo = "text") %>%
  add_markers(alpha = 0.1,text=paste(gap$country,gap$pop), color = I("black")) %>%
  add_markers(
    data = gapKey, 
    frame = ~year, 
    ids = ~country, 
    text=paste(gap$country,gap$pop),
    color = I("red")
  ) %>%
  layout(xaxis = list(type = "log"))

p2 <- plot_ly(gap, x = ~gdpPercap, y = ~lifeExp, 
              text = ~country, hoverinfo = "text") %>%
  add_markers(color = I("black"), alpha = 0.1) %>%
  add_markers(
    data = gapKey, 
    frame = ~year, 
    ids = ~country, 
    color = I("red")
  ) %>%
  layout(xaxis = list(type = "log"))

subplot(p1, p2, nrows = 1, widths = c(0.3, 0.7), titleX = TRUE) %>%
  hide_legend() %>%
  animation_opts(1000, redraw = FALSE) %>%
  layout(hovermode = "y", margin = list(l = 100)) %>%
  highlight(
    "plotly_selected", 
    color = "blue", 
    opacityDim = 1, 
    hoverinfo = "none"
  )
```



