---
title: 'Interactive Data Visualization with plotly in R '
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: false
    toc_float: true
    toc_depth: 3
    css: doc.css
    df_print: paged
  pdf_document: default
---


```{r eval=FALSE,warning = FALSE, message = FALSE}
# used libraries
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("plotly")) install.packages("plotly")
# R package DT provides an R interface to the JavaScript library DataTables.
if (!require("DT")) install.packages("DT")
# geom_mark_hull
if (!require("concaveman")) install.packages("concaveman")
if (!require("ggforce")) install.packages("ggforce")


# more libraries
if (!require("listviewer")) install.packages("listviewer")
if (!require("lubridate")) install.packages("lubridate")
if (!require("forcats")) install.packages("forcats")
if (!require("maps")) install.packages("maps")
if (!require("mvtnorm")) install.packages("mvtnorm")
if (!require("gapminder")) install.packages("gapminder")
if (!require("hexbin")) install.packages("hexbin")
if (!require("Hmisc")) install.packages("Hmisc")

```
```{r setup, include=FALSE,warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
library(ggplot2)
library(plotly)
library(ggforce)
library(dplyr)
library(lubridate)
```

# plotly - Documentation

* Documentation:
  * [https://plotly.com/r/](https://plotly.com/r/)
* Book:
  * Interactive web-based data visualization with R, plotly, and shiny
by Carson Sievert [https://plotly-r.com/](https://plotly-r.com/)

This document is essentially a brief notes of the book above.

# Little Bit of Low-Level Details

## Figure as List: list -> json -> plotly.js

```{r}
library(plotly)
fig1 <- list(
  data = list(
    list(
      x = c(1, 2, 3),
      y = c(1, 3, 2),
      type = 'bar'
    )
  ),
  layout = list(
    title = 'A Figure Specified By R List',
    plot_bgcolor='#e5ecf6', 
         xaxis = list( 
           zerolinecolor = '#ffff', 
           zerolinewidth = 2, 
           gridcolor = 'ffff'), 
         yaxis = list( 
           zerolinecolor = '#ffff', 
           zerolinewidth = 2, 
           gridcolor = 'ffff')
  )
)
# To display the figure defined by this list, use the plotly_build function
plotly_build(fig1)
```
```{r}
library(plotly) 

fig2 <- plot_ly(x = c(1, 2, 3), y = c(1, 3, 2), type = 'bar')%>% 
  layout(title = 'A Plotly Figure',
         plot_bgcolor='#e5ecf6', 
         xaxis = list( 
           zerolinecolor = '#ffff', 
           zerolinewidth = 2, 
           gridcolor = 'ffff'), 
         yaxis = list( 
           zerolinecolor = '#ffff', 
           zerolinewidth = 2, 
           gridcolor = 'ffff')) 
fig2
```
```{r}
str(fig2, max.level = 2)
```

```{r}
plotly_json(fig2)
```
![plotly.R to plotly.js](https://plotly-r.com/images/printing.svg)


* In principle you can update plotly list object to achive a particular effect
  * Typically we don't whan to use it  
* plotly.R has many good routines to make our life easier


# Overview

The Figure Data Structure (plotly object attributes):

* data or traces - contains data and ways how to plot it
* layout - control positioning and configuration of non-data-related parts 
  of the figure (dimentions, margins, title, legend, ..)
* frames - used for animated plots


Creating plotly object

* initialize a plotly object with `plot_ly()`/`plot_geo()`/`plot_mapbox()`
* convert ggplot (using `ggplotly()`)
* convert list (using `plotly_build()`)


plot_ly():

*  ‘direct’ interface to plotly.js
*  additional abstractions to help reduce typing
*  Similar to ggplot2 plotly also implement Grammar of Graphics


## Into to plot_ly

```{r}
# load the plotly R package
library(plotly)

# load the diamonds dataset from the ggplot2 package
data(diamonds, package = "ggplot2")
#knitr::kable(diamonds)
diamonds
```

```{r}
# create three visualizations of the diamonds dataset
plot_ly(diamonds, x = ~cut)
plot_ly(diamonds, x = ~cut, y = ~clarity)

# using Accent palettes from RColorBrewer try Set1, Paired and others
plot_ly(diamonds, x = ~cut, color = ~clarity, colors = "Accent", type="histogram")
```

```{r}
# doesn't produce black bars, 'black' is value here
plot_ly(diamonds, x = ~cut, color = "black")
# produces red bars with black outline
plot_ly(
  diamonds, 
  x = ~cut, 
  color = I("red"), 
  stroke = I("black"), 
  span = I(2)
)
```

> Use `I(<value>)` to specify color, mark and similar
>


## Chaining with %>% 

Most function from plotly takes a plotly object as first argument and thus can participate in %>% chaining.

```{r}
layout(
  plot_ly(diamonds, x = ~cut),
  title = "My beatiful histogram"
)
```

%>% often produce more human readable code

```{r}
diamonds %>%
  plot_ly(x = ~cut) %>%
  layout(title = "My beatiful histogram")
```

```{r}
diamonds %>%
  plot_ly() %>% 
  add_histogram(x = ~cut)
```



```{r}
diamonds %>%
  dplyr::count(cut) %>%
  plot_ly() %>% 
  add_bars(x = ~cut, y = ~n)
```


What is the difference between two above plots?

Calculations done in different places (it can be important for speed and interactivity).

add_*() functions:
* add_histogram2d(), add_contour(), add_boxplot()

plot_ly():
* Arguments specified in plot_ly() are global, meaning that any downstream add_*() functions inherit these arguments (unless inherit = FALSE).
* manipulation verbs from the dplyr package may be used to transform the data underlying a plotly object.
* use the plotly_data() to get data

```{r}
library(dplyr)

diamonds %>%
  plot_ly(x = ~cut) %>% 
  add_histogram() %>%
  group_by(cut) %>%
  summarise(n = n()) %>%
  add_text(
    text = ~scales::comma(n), y = ~n, 
    textposition = "top middle", 
    cliponaxis = FALSE
  )
```

## Into to ggplotly()

* Converts ggplot graphics to plotly.
* plotly does not have every possible plot in plot_ly(), so ggplotly extends
  interactivity to large set of ggplots.
* Large number of plots converts properly.

```{r}
p <- ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_hex(bins = 100)
ggplotly(p)
```

```{r}
p <- ggplot(diamonds, aes(x = log(price), color = clarity)) + 
    geom_freqpoly()
ggplotly(p)
```

```{r}
p <- ggplot(diamonds, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + 
    facet_wrap(~cut)
ggplotly(p)
```
```{r}
m <- lm(log(price) ~ log(carat), data = diamonds)
diamonds <- modelr::add_residuals(diamonds, m)
p <- ggplot(diamonds, aes(x = clarity, y = resid, color = clarity)) +
    ggforce::geom_sina(alpha = 0.1) + 
    stat_summary(fun.data = "mean_cl_boot", color = "black") +
    facet_wrap(~cut)

# WebGL is a lot more efficient at rendering lots of points
toWebGL(ggplotly(p))
```

# Scattered foundations

Some scatter type add_* routines:
* add_markers()
* add_lines()
* add_paths()
* add_segments()
* add_ribbons()
* add_area()
* add_polygons()


```{r}
library(plotly)
data(economics, package = "ggplot2")
economics
```

```{r}
# show difference between paths and lines
p <- economics[sample(nrow(economics)),] %>%
  plot_ly(x = ~date, y = ~psavert)

add_paths(p)
add_lines(p)
add_lines(p) %>% rangeslider()
```
```{r}
library(lubridate)
econ <- economics %>%
  mutate(yr = year(date), mnth = month(date))

# One trace (more performant, but less interactive) NOTE: group_by use
econ %>%
  group_by(yr) %>%
  plot_ly(x = ~mnth, y = ~uempmed) %>%
  add_lines()

# Multiple traces (less performant, but more interactive)
plot_ly(econ, x = ~mnth, y = ~uempmed) %>%
  add_lines(color = ~ordered(yr))
  
# The split argument guarantees one trace per group level (regardless 
# of the variable type). This is useful if you want a consistent
# visual property over multiple traces 
plot_ly(econ, x = ~mnth, y = ~uempmed) %>%
   add_lines(split = ~yr, color = I("black"))

#three above methods vary in interactivity and scaling/performance
```
```{r}
# more raw plotly.js style, only add_trace
set.seed(99)
plot_ly() %>%
 add_trace(
   type = "scatter",
   mode = "markers+lines+text",
   x = 4:6, 
   y = 4:6,
   text = replicate(3, "Here is the text 🙌"),
   textposition = "right",
   hoverinfo = "text",
   textfont = list(family = "Roboto Condensed", size = 16)
 ) %>%
 layout(xaxis = list(range = c(3, 8)))

```

```{r}
schema()
```

```{r}
# add_markers adds markers/points
# alpha color can help with overcrowding
# too many points can be compute intensive, try toWebGL
# eventually you will need to use 2d histogram

subplot(
  plot_ly(mpg, x = ~cty, y = ~hwy, name = "default"),
  plot_ly(mpg, x = ~cty, y = ~hwy) %>% 
    add_markers(alpha = 0.2, name = "alpha")
)
```

use `suplot()` for placing multiple plots on same figure

```{r}
# use help and make it share X and Y axes

subplot(
  plot_ly(mpg, x = ~cty, y = ~hwy, name = "default"),
  plot_ly(mpg, x = ~cty, y = ~hwy) %>% 
    add_markers(alpha = 0.2, name = "alpha"),
  shareX=TRUE, shareY=T
)
```

Discrete vs continuous color scale.
```{r}
# discrete variable to color produces one trace per category,
# numeric variable to color produces one trace
# colorbar() - ustomize the appearance of scale bar
p <- plot_ly(mpg, x = ~cty, y = ~hwy, alpha = 0.5)
subplot(
  add_markers(p, color = ~cyl, showlegend = FALSE) %>% 
    colorbar(title = "Viridis"),
  add_markers(p, color = ~factor(cyl))
)
```

# Specifying color palette.

Palettes types:
* Sequential
  * gradual change from one color to another
  * typically light color to dark color
  * Good for continuous data, differentiate min and max
* Diverging
  * Often gradual change between three colors
  * Good for continuous data, differentiate min, medium and max)
* Qualitative
  * discrete color change, distinct colors between different values
  * Good for categories 
  
```{r fig.height=7,fig.width=7}
library(RColorBrewer)
par(mar=c(3,4,2,2))
display.brewer.all()
```
```{r}
subplot(
  add_markers(p, color = ~cyl, colors = c("red", "#000000", "blue")) %>%
    colorbar(title = "plotly"),
  add_markers(p, color = ~cyl, colors = "Spectral") %>% # RColorBrewer::brewer.pal.info for more
    colorbar(title = "RColorBrewer::Spectral"),
  add_markers(p, color = ~cyl, colors = viridisLite::inferno(10)) %>% 
    colorbar(title = "viridisLite::Inferno"),
  add_markers(p, color = ~cyl, colors = colorRamp(c("red", "white", "blue"))) %>% 
    colorbar(title = "colorRamp"),
  nrows = 2
) %>% hide_legend()
```

```{r}
subplot(
  # plotly
  add_markers(p, color = ~factor(cyl), colors = c("red", "#000000", "blue","green"), 
              legendgroup = '1'),
  # plotly by values
  add_markers(p, color = ~factor(cyl), colors = c('4'="red", '5'="black", '6'="blue", '8'="green"), 
              legendgroup = '2'),
  # RColorBrewer::Set1
  add_markers(p, color = ~factor(cyl), colors = "Set1", 
              legendgroup = '3'),
  # colorRamp or other continious maps
  add_markers(p, color = ~factor(cyl), colors = colorRamp(c("red", "white", "blue")), 
              legendgroup = '4'),
  nrows = 2
)
# one legend per whole plot at this time, legendgroup allows grouping and some padding
```


```{r}
subplot(
  add_markers(p, color = ~cyl, colors = c("red", "#000000", "blue")) %>%
    colorbar(title = "plotly"),
  add_markers(p, color = ~cyl, colors = "Spectral") %>% # RColorBrewer::brewer.pal.info for more
    colorbar(title = "RColorBrewer::Spectral"),
  add_markers(p, color = ~cyl, colors = viridisLite::inferno(10)) %>% 
    colorbar(title = "viridisLite::Inferno"),
  add_markers(p, color = ~cyl, colors = colorRamp(c("red", "white", "blue"))) %>% 
    colorbar(title = "colorRamp"),
  nrows = 2
) %>% hide_legend()
```


Symbols
```{r}
p <- plot_ly(mpg, x = ~cty, y = ~hwy, alpha = 0.3) 
subplot(
  add_markers(p, symbol = ~cyl, name = "A single trace"),
  add_markers(p, symbol = ~factor(cyl), color = I("black")),
  add_markers(
    p, color = I("black"),
    symbol = ~factor(cyl), 
    symbols = c("triangle-up", "diamond", "circle")
  ),
  add_markers(p, size = ~cyl, sizes = c(1, 500), name = "custom"),
  nrows = 2
)
# stroke, span, size also can be changed
```
```{r}
# add_segments
mpg %>%
  group_by(model) %>%
  summarise(c = mean(cty), h = mean(hwy)) %>%
  mutate(model = forcats::fct_reorder(model, c)) %>% # forcats::fct_reorder for factor reorder
  plot_ly() %>%
  add_segments(
    x = ~c, y = ~model,
    xend = ~h, yend = ~model, 
    color = I("gray"), showlegend = FALSE
  ) %>%
  add_markers(
    x = ~c, y = ~model, 
    color = I("blue"), 
    name = "mpg city"
  ) %>%
  add_markers(
    x = ~h, y = ~model, 
    color = I("red"),
    name  = "mpg highway"
  ) %>%
  layout(xaxis = list(title = "Miles per gallon"))
```

```{r}
# leveraging density function to plot kernel dencity estimation
kerns <- c("gaussian", "epanechnikov", "rectangular", 
          "triangular", "biweight", "cosine", "optcosine")
p <- plot_ly()
for (k in kerns) {
  d <- density(economics$pce, kernel = k, na.rm = TRUE)
  p <- add_lines(p, x = d$x, y = d$y, name = k)
}
p
# or
ggplotly(ggplot(economics, aes(x=pce))+geom_density())
```

see more <https://plotly-r.com/scatter-traces.html>



