---
title: 'L01. Interactive Data Visualization with plotly in R: S01. Motivation'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: false
    toc_float: true
    toc_depth: 3
    css: doc.css
    df_print: paged
  pdf_document: default
always_allow_html: true
---


```{r eval=FALSE,warning = FALSE, message = FALSE}
# used libraries
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("plotly")) install.packages("plotly")
# R package DT provides an R interface to the JavaScript library DataTables.
if (!require("DT")) install.packages("DT")
# geom_mark_hull
if (!require("concaveman")) install.packages("concaveman")
if (!require("ggforce")) install.packages("ggforce")
```
```{r setup, include=FALSE,warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotly)
library(ggforce)
```

# plotly - Documentation

* Documentation:
  * [https://plotly.com/r/](https://plotly.com/r/)
* Book:
  * Interactive web-based data visualization with R, plotly, and shiny
by Carson Sievert [https://plotly-r.com/](https://plotly-r.com/)

This document is essentially a brief notes of the book above.

# Motivation

* Why interactive data visualization (with plotly)?
* What are the benifits over ggplot or basic plots?

```{r warning = FALSE, message = FALSE}
library(ggplot2)
mpg
```
Exploratory Data Analysis with Static plots
```{r}
library(ggplot2)
ggplot(mpg, aes(displ, hwy)) + geom_point()
```

What are max hwy? 'Outliers`?


```{r}
library(plotly)
p <- ggplot(mpg, aes(displ, hwy)) + geom_point()
ggplotly(p)
```



What are those models:
```{r}
library(plotly)
m <- highlight_key(mpg)
p <- ggplot(m, aes(displ, hwy)) + geom_point()
gg <- highlight(ggplotly(p), "plotly_selected")
crosstalk::bscols(gg, DT::datatable(m))
```


Final figure:
```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  ggforce::geom_mark_hull(aes(filter = model == "corvette", label = model)) +
  labs(
    title = "Fuel economy from 1999 to 2008 for 38 car models",
    caption = "Source: https://fueleconomy.gov/",
    x = "Engine Displacement", 
    y = "Miles Per Gallon"
  )
```



Interactions:
```{r}
library(plotly)
p <- ggplot(mpg, aes(displ, hwy,color=model)) + geom_point()
ggplotly(p)
```
* hover over data-point - value
* click-and-drag (horizontally, vertically, arbitrary) - zoom
* click-and-drag axis - move alone axis
* click-and-drag on axis corner - move axis corner value
* shift-click-drag - move data view (pan)
* double-click - restore view
* click on legend item - show/hide items
* double click on legend item - show/hide single item



