---
title: 'Interactive Data Visualization with plotly in R '
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    css: doc.css
    df_print: paged
  pdf_document: default
---


```{r eval=FALSE,warning = FALSE, message = FALSE}
# used new libraries
if (!require("shiny")) install.packages("shiny")
if (!require("readr")) install.packages("readr")
if (!require("purrr")) install.packages("purrr")

```

```{r setup, include=FALSE,warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
# library(ggplot2)
# library(plotly)
# library(ggforce)
# library(dplyr)
# library(lubridate)
```

# Rshiny 

Shiny – an R package for creating reactive web applications entirely in R. 

Shiny’s **reactive programming model** allows R programmers to build upon their existing R knowledge and create data-driven web applications without any prior web programming experience. 


# Rshiny - Documentation and Resources

* Documentation:
  * Documentation: <https://shiny.rstudio.com/articles/>
  * Reference: <https://shiny.rstudio.com/reference/shiny/>
  * Official Tutorial: <https://shiny.rstudio.com/tutorial/>
    * <https://github.com/rstudio-education/shiny.rstudio.com-tutorial>
  * Examples: <https://shiny.rstudio.com/gallery/>
* Book:
  * Mastering Shiny<https://mastering-shiny.org/>
  * Interactive web-based data visualization with R, plotly, and shiny
by Carson Sievert <https://plotly-r.com/>

This document is essentially a brief notes of the book above.

# Example 1

```{r}
# dataset txhousing - houses prices in Texas cities
# txhousing dataframe comes with plotly library
library(plotly)

data(txhousing)
txhousing
```

With ggplot2 for plotting
```{r eval=FALSE,warning = FALSE, message = FALSE}
library(shiny)
library(ggplot2)
library(dplyr)

# User Interface Description
# - defines how inputs and output widgets are displayed on the page
ui <- fluidPage(
  selectInput(
    inputId = "cities", 
    label = "Select a city", 
    choices = unique(txhousing$city), 
    selected = "Abilene",
    multiple = TRUE
  ),
  plotOutput(outputId = 'ggplot1')
)

# The server function
# - defines a mapping from input values to output widgets
server <- function(input, output, session) {
  # note that element of output matches outputId in ui
  output$ggplot1 <- renderPlot({ 
    # note that element of input matches inputId in ui
    ggplot(txhousing %>% filter(city %in% input$cities),
           aes(x=date, y=median, color=city)) + geom_line()
    # what is that syntax and what does it return?
  })
}

# Start shiny web-server
shinyApp(ui, server)
```
First argument, in **reactive function** (`render*`()), `{}`, is expression which possibly contains **reactive values**. This reactive expression will be called automatically on change of reactive values!

Reactive values notify reactive function object that it value changes, in other ward that the old value is invalid now (thus it is also called invalidation).


With plotly for plotting
```{r eval=FALSE,warning = FALSE, message = FALSE}
library(shiny)
library(plotly)


# User Interface Description
# - defines how inputs and output widgets are displayed on the page
ui <- fluidPage(
  selectizeInput(
    inputId = "cities", 
    label = "Select a city", 
    choices = unique(txhousing$city), 
    selected = "Abilene",
    multiple = TRUE
  ),
  plotlyOutput(outputId = "plotly1")
)

# The server function
# - defines a mapping from input values to output widgets
server <- function(input, output, session) {
  # note that element of output matches outputId in ui
  output$plotly1 <- renderPlotly({
    # note that element of input matches inputId in ui
    plot_ly(txhousing, x = ~date, y = ~median) %>%
      filter(city %in% input$cities) %>%
      group_by(city) %>%
      add_lines()
  })
}

# Start shiny web-server
shinyApp(ui, server)
```

Standard Shiny input widgets:

* selectInput()/selectizeInput() for dropdown menus.
* numericInput() for a single number.
* sliderInput() for a numeric range.
* textInput() for a character string.
* dateInput() for a single date.
* dateRangeInput() for a range of dates.
* fileInput() for uploading files.
* checkboxInput()/checkboxGroupInput()/radioButtons() for choosing a list of options.

Standard Output widgets (ui function / rander function):

* plotOutput() / renderPlot()
  * static base and ggplot
* tableOutput() / renderTable() ; datatableOutput() / renderDataTable()
  * table
* verbatimTextOutput() / renderPrint() ; textOutput()/renderText() ; 
  * text output
* imageOutput() / renderImage()
  * image output

More output widgets:
* plotlyOutput() / renderPlotly()
  * plotly
* leafletOutput() / renderLeaflet()
  * maps

Output widgets can be input too (plots/ggplots/plotly/DT/leaflet)

Almost all plotly events can be be an input

```{r eval=FALSE,warning = FALSE, message = FALSE}
plotly_example("shiny", "event_data")
```

Add events handling to TXHousing


## Edit events

Edit events:  you can directly manipulate annotations, title, shapes by adding
config(p, editable = TRUE). This editing generates plotly_relayout event. By listening to it we can generate a responce to edit.

```{r eval=FALSE,warning = FALSE, message = FALSE}
# Ex02_drag_markers.R
# the effect of high leverage points on a simple linear model
plotly_example("shiny", "drag_markers")
```
Also check "plotly_restyle" events.

## Scoping events

When managing multiple views that communicate with one another, you’ll need to be aware of which views are a source of interaction and which are a target (a view can be both, at once!). 

The event_data() function provides a source argument to help refine which view(s) serve as the source of an event.

The source argument takes a string ID, and when that ID matches the source of a plot_ly()/ggplotly() graph, then the event_data() is “scoped” to that view. 


```{r eval=FALSE,warning = FALSE, message = FALSE}
library(shiny)

# cache computation of the correlation matrix
correlation <- round(cor(mtcars), 3)

ui <- fluidPage(
  plotlyOutput("heat"),
  plotlyOutput("scatterplot")
)

server <- function(input, output, session) {
  
  output$heat <- renderPlotly({
    plot_ly(source = "heat_plot") %>%
      add_heatmap(
        x = names(mtcars), 
        y = names(mtcars), 
        z = correlation
      )
  })
  
  output$scatterplot <- renderPlotly({
    # if there is no click data, render nothing!
    clickData <- event_data("plotly_click", source = "heat_plot")
    if (is.null(clickData)) return(NULL)
    
    # Obtain the clicked x/y variables and fit linear model
    vars <- c(clickData[["x"]], clickData[["y"]])
    d <- setNames(mtcars[vars], c("x", "y"))
    yhat <- fitted(lm(y ~ x, data = d))
    
    # scatterplot with fitted line
    plot_ly(d, x = ~x) %>%
      add_markers(y = ~y) %>%
      add_lines(y = ~yhat) %>%
      layout(
        xaxis = list(title = clickData[["x"]]), 
        yaxis = list(title = clickData[["y"]]), 
        showlegend = FALSE
      )
  })
  
}

shinyApp(ui, server)
```

```{r eval=FALSE,warning = FALSE, message = FALSE}
plotly_example("shiny", "event_data_persist")
```

```{r eval=FALSE,warning = FALSE, message = FALSE}
# 01-kmeans-app

palette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
          "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999"))

library(shiny)

ui <- fluidPage(
  headerPanel('Iris k-means clustering'),
  sidebarPanel(
    selectInput('xcol', 'X Variable', names(iris)),
    selectInput('ycol', 'Y Variable', names(iris),
                selected = names(iris)[[2]]),
    numericInput('clusters', 'Cluster count', 3,
                 min = 1, max = 9)
  ),
  mainPanel(
    plotOutput('plot1')
  )
)

server <- function(input, output) {
  
  selectedData <- reactive({
    iris[, c(input$xcol, input$ycol)]
  })
  
  clusters <- reactive({
    kmeans(selectedData(), input$clusters)
  })
  
  output$plot1 <- renderPlot({
    par(mar = c(5.1, 4.1, 0, 1))
    plot(selectedData(),
         col = clusters()$cluster,
         pch = 20, cex = 3)
    points(clusters()$centers, pch = 4, cex = 4, lwd = 4)
  })
  
}

shinyApp(ui = ui, server = server)
```



