---
title: "Ch12-unsup-lab"
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r eval=FALSE,warning = FALSE, message = FALSE}
# used new libraries
if (!require("ISLR2")) install.packages("ISLR2")
if (!require("cluster")) install.packages("cluster")

```

```{r setup, include=FALSE,warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
library(plotly)
library(cluster)
```


# NCI60 Data Example


Unsupervised techniques are often used in the analysis of genomic data. In particular, PCA and hierarchical clustering are popular tools.
 We  illustrate these techniques on the `NCI` cancer cell line microarray data, which consists of $6{,}830$ gene expression measurements on $64$ cancer cell lines.

```{r chunk41}
library(ISLR2)
nci.labs <- NCI60$labs
nci.data <- NCI60$data
```

Each cell line is labeled with a cancer type, given in `nci.labs`. We do not make use of the cancer types in performing PCA and clustering, as these are unsupervised techniques. But
after performing PCA and clustering, we will
check to see the extent to which these cancer types agree with the results of these unsupervised techniques.

The data has $64$ rows and $6{,}830$ columns.

```{r chunk42}
dim(nci.data)
```


We begin by examining the cancer types for the cell lines.

```{r chunk43}
nci.labs[1:4]
table(nci.labs)
```


### PCA on the NCI60 Data

We first perform PCA on the data after scaling the variables (genes) to have standard deviation one, although one could reasonably argue that it is better not to scale the genes.

```{r chunk44}
pr.out <- prcomp(nci.data, scale = TRUE)
```

We now  plot the first few principal component score vectors, in order to visualize the data. The observations (cell lines) corresponding to a given cancer type will be plotted in the same color, so that we can see to what extent the observations within a cancer type are similar to each other. We first create a simple function that assigns a distinct color to each element of a numeric vector.
The function will be used to assign a  color to each of the $64$ cell lines, based on the cancer type to which it corresponds.

```{r chunk45}
Cols <- function(vec) {
   cols <- rainbow(length(unique(vec)))
   return(cols[as.numeric(as.factor(vec))])
 }
```

Note that the `rainbow()` function takes as its argument a positive integer, and returns a vector containing that number of distinct colors.  We now can plot the principal component score vectors.

```{r chunk46}
par(mfrow = c(1, 2))
plot(pr.out$x[, 1:2], col = Cols(nci.labs), pch = 19,
    xlab = "Z1", ylab = "Z2")
plot(pr.out$x[, c(1, 3)], col = Cols(nci.labs), pch = 19,
    xlab = "Z1", ylab = "Z3")
```

The resulting  plots are shown in Figure 12.17. On the whole, cell lines corresponding to a single cancer type do tend to have similar values on the first few
principal component score vectors. This indicates that cell lines from the same cancer type tend to have pretty similar gene expression levels.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{Figures/Chapter12/12_17.pdf} 
On the whole, observations belonging to a single cancer type tend to lie near each other
in this low-dimensional space. It would not have been possible to visualize the data without using a dimension reduction method such as PCA, since based on the full data set there are
 $6{,}830 \choose 2$ possible scatterplots, none of which would have
been particularly informative.}
\end{figure}

We can obtain a summary of the proportion of variance explained (PVE) of the first few principal components using the `summary()` method for a `prcomp` object (we have truncated the printout):

```{r chunk47}
summary(pr.out)
```

Using the `plot()` function, we can also plot the variance explained by the first few principal components.

```{r chunk48}
plot(pr.out)
```

Note that the height of each bar in the bar plot is given by squaring the corresponding element of `pr.out$sdev`.
However, it is more informative to plot the PVE of each principal component (i.e. a scree plot) and the cumulative PVE of each principal component. This can be done with just a little work.

```{r chunk49}
pve <- 100 * pr.out$sdev^2 / sum(pr.out$sdev^2)
par(mfrow = c(1, 2))
plot(pve,  type = "o", ylab = "PVE",
    xlab = "Principal Component", col = "blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE",
    xlab = "Principal Component", col = "brown3")
```

(Note that the elements of `pve` can also be computed directly from the summary, `summary(pr.out)$importance[2, ]`, and the elements of
`cumsum(pve)` are given by
`summary(pr.out)$importance[3, ]`.)
The resulting plots are shown in Figure 12.18.
\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{Figures/Chapter12/12_18.pdf}
*Right:* the cumulative PVE of the principal components is shown. Together, all principal components explain $100\,\%$ of the variance.}
\end{figure}
We see that together,
the first seven principal components explain around $40\,\%$ of the variance in the data. This is not a huge
amount of the variance. However, looking at the scree plot, we see that while each of the first seven principal components explain a substantial amount of  variance, there
is a marked decrease in the variance explained by further principal components. That is, there is an *elbow*   in the plot after approximately the seventh principal component.
This suggests that there may be little benefit to examining more than seven or so principal components (though even examining seven principal components may be difficult).



### Clustering the Observations of the NCI60 Data

We now proceed to hierarchically cluster the cell lines in the `NCI` data, with the goal of finding out whether or not the observations cluster into distinct types of cancer. To begin, we standardize the variables to have mean
 zero and standard deviation one. As mentioned earlier, this step is optional and should be performed only if we want each gene to be on the same *scale*.

```{r chunk50}
sd.data <- scale(nci.data)
```

We now perform hierarchical clustering of the observations using complete, single, and average linkage. Euclidean distance is used as the dissimilarity measure.

```{r chunk51}
par(mfrow = c(1, 3))
data.dist <- dist(sd.data)
plot(hclust(data.dist), xlab = "", sub = "", ylab = "",
    labels = nci.labs, main = "Complete Linkage")
plot(hclust(data.dist, method = "average"),
    labels = nci.labs, main = "Average Linkage",
    xlab = "", sub = "", ylab = "")
plot(hclust(data.dist, method = "single"),
    labels = nci.labs,  main = "Single Linkage",
    xlab = "", sub = "", ylab = "")
```

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{Figures/Chapter12/12_19.pdf}
as the dissimilarity measure. Complete and average linkage tend to yield
evenly sized clusters whereas single linkage tends to yield extended clusters to which single leaves are fused one by one.}
\end{figure}
The results are shown in Figure 12.19.
We see that the choice of linkage certainly  does affect the results obtained. Typically, single linkage will tend to yield *trailing* clusters: very large clusters onto which
 individual observations attach  one-by-one. On the other hand, complete and average linkage tend to yield more balanced, attractive clusters. For this reason, complete
and average linkage are generally preferred to single linkage.
Clearly cell lines within a single cancer type do tend to cluster together, although the clustering is not perfect. We will use complete linkage hierarchical clustering for the analysis that follows.


We can cut the dendrogram at the height that will yield a particular number of clusters, say four:

```{r chunk52}
hc.out <- hclust(dist(sd.data))
hc.clusters <- cutree(hc.out, 4)
table(hc.clusters, nci.labs)
```

There are some clear patterns. All the leukemia cell lines fall in cluster $3$, while the breast cancer cell lines are spread out over three different clusters.  We can plot the cut on the dendrogram that produces these four clusters:

```{r chunk53}
par(mfrow = c(1, 1))
plot(hc.out, labels = nci.labs)
abline(h = 139, col = "red")
```

The `abline()` function draws a straight line on top of any existing plot in~`R`. The argument `h = 139` plots a horizontal line at height $139$ on the
dendrogram; this is the height that results in four distinct clusters. It is easy to verify that the resulting clusters are the same as the ones we obtained using `cutree(hc.out, 4)`.


Printing the output of `hclust` gives a useful brief summary of the object:

```{r chunk54}
hc.out
```


We claimed earlier in Section 12.4.2 that $K$-means clustering and hierarchical clustering with the dendrogram cut to obtain the same number of clusters can yield very different results.
How do these `NCI` hierarchical clustering results compare to what we  get if we perform $K$-means clustering with $K=4$?

```{r chunk55}
set.seed(2)
km.out <- kmeans(sd.data, 4, nstart = 20)
km.clusters <- km.out$cluster
table(km.clusters, hc.clusters)
```

We see that the four clusters obtained using hierarchical clustering and $K$-means clustering  are somewhat different. Cluster~$4$ in $K$-means clustering is identical to cluster~$3$
in hierarchical clustering. However, the other clusters differ: for instance, cluster~$2$ in $K$-means clustering contains a portion of the observations assigned to
cluster 1 by hierarchical clustering, as well as all of the observations assigned to cluster~$2$ by hierarchical clustering.

Rather than performing hierarchical clustering on the entire data matrix, we can simply perform hierarchical clustering on the first few principal component score vectors,
as follows:

```{r chunk56}
hc.out <- hclust(dist(pr.out$x[, 1:5]))
plot(hc.out, labels = nci.labs,
    main = "Hier. Clust. on First Five Score Vectors")
table(cutree(hc.out, 4), nci.labs)
```

 Not surprisingly, these results are different from the ones that we
 obtained when we performed hierarchical clustering on the full data
 set. Sometimes performing clustering on the first few principal
 component score vectors can give better results than performing
 clustering on the full data.  In this situation, we might view the principal
component step as one of denoising the data.
We could also perform $K$-means
 clustering on the first few principal component score vectors rather
 than the full data set.






