---
title: "Ethereum_Dogecoin_value_forecasting"
output: html_document
date: "2022-12-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r setup, warning=F, message=F,echo=F}
# Importing the libraries
library(ggplot2)
library(dplyr)
library(plotly)
library(hrbrthemes)
library(forecast)
library(MASS) 
library(tseries)
```

```{r}
# Loading the dataset for Ethereum
ethereum <- read.csv("coin_Ethereum.csv",header=T)
ethereum$Date <- as.Date(ethereum$Date)

# # Loading the dataset for Dogecoin
dogecoin <- read.csv("coin_Dogecoin.csv",header=T)
dogecoin$Date <- as.Date(dogecoin$Date)
```

```{r}
# Plotting the time series for Ethereum
plot_eth_ts <- ethereum %>%
  ggplot( aes(x=Date, y=Close)) +
    geom_area(fill="#69b3a2", alpha=0.5) +
    geom_line(color="#69b3a2") +
    ylab("Ethereum Closing price ($)") +
    theme_classic()

plot_eth_ts <- ggplotly(plot_eth_ts)
plot_eth_ts
```

Plotting Closing prices of Ethereum with best fit line
```{r}
# plotting with a LOESS line
par(mfrow = c(1, 1))
ggplot(ethereum, aes(x = Date, y = Close, group = 1)) +
  geom_line() +
  xlab("Date") +
  ylab("USD") +
  ggtitle("Closing Price of Ethereum") +
  geom_smooth(method = "loess", level = 0)
```

Plotting Closing prices of Dogecoin with best fit line
```{r}
# plotting with a LOESS line
par(mfrow = c(1, 1))
ggplot(dogecoin, aes(x = Date, y = Close, group = 1)) +
  geom_line() +
  xlab("Date") +
  ylab("USD") +
  ggtitle("Closing Price of Dogecoin") +
  geom_smooth(method = "loess", level = 0)
```

```{r}
# Plotting the time series for DogeCoin
plot_doge_ts <- dogecoin %>%
  ggplot( aes(x=Date, y=Close)) +
    geom_area(fill="#69b3a2", alpha=0.5) +
    geom_line(color="#69b3a2") +
    ylab("Dogecoin Closing price ($)") +
    theme_classic()

plot_doge_ts <- ggplotly(plot_doge_ts)
plot_doge_ts
```

Creating a time series object for the closing pices of Ethereum and Dogecoin cryptocurrencies
```{r}
close_eth <- as.ts(ethereum$Close)
close_doge <- as.ts(dogecoin$Close)
```

ACF and PACF plots for Ethereum
```{r}
ggtsdisplay(close_eth)
```

ACF and PACF plots for Dogecoin
```{r}
ggtsdisplay(close_doge)
```
ACF and PACF plots Ethereum plotly
```{r}
library(BETS)
corrgram(close_eth, lag.max = 35, type = "correlation", mode = "simple",
  ci = 0.95, style = "plotly", knit = F)
corrgram(close_eth, lag.max = 35, type = "partial", mode = "simple",
  ci = 0.95, style = "plotly", knit = F)
```

ACF and PACF plots Dogecoin plotly
```{r}
corrgram(close_doge, lag.max = 35, type = "correlation", mode = "simple",
  ci = 0.95, style = "plotly", knit = F)
corrgram(close_doge, lag.max = 35, type = "partial", mode = "simple",
  ci = 0.95, style = "plotly", knit = F)
```

Box-Cox tranformation on Ethereum
```{r}
par(mfrow = c(1, 1))
# library(MASS) is loaded
boxcox_eth_transf <- boxcox(close_eth ~ as.numeric(1:length(close_eth)))
```

Box-Cox tranformation on Dogecoin
```{r}
par(mfrow = c(1, 1))
# library(MASS) is loaded
boxcox_doge_transf <- boxcox(close_doge ~ as.numeric(1:length(close_doge)))
```

Computing value of lambda fo which maximum log-likelihood is achieved for Ethereum
```{r}
lambda_eth <- boxcox_eth_transf$x[which(boxcox_eth_transf$y == max(boxcox_eth_transf$y))]
lambda_eth
```

Computing value of lambda fo which maximum log-likelihood is achieved for Dogecoin
```{r}
lambda_doge <- boxcox_doge_transf$x[which(boxcox_doge_transf$y == max(boxcox_doge_transf$y))]
lambda_doge
```

Transforming the Ethereum data and plotting the transformed data with the line of best fit
```{r}
# transforming the data
ethereum$close_transf_eth <- log(close_eth)

# new time series plot
par(mfrow = c(1, 1))
ggplot(ethereum, aes(x = Date, y = close_transf_eth, group = 1)) +
  geom_line() +
  xlab("Date") +
  ylab("USD (after log transformation)") +
  ggtitle("Series after Box-Cox (Ethereum)") +
  geom_smooth(method = "loess", level = 0)
```

Transforming the Dogecoin data and plotting the transformed data with the line of best fit
```{r}
# transforming the data
dogecoin$close_transf_doge <- log(close_doge)

# new time series plot
par(mfrow = c(1, 1))
ggplot(dogecoin, aes(x = Date, y = close_transf_doge, group = 1)) +
  geom_line() +
  xlab("Date") +
  ylab("USD (after log transformation)") +
  ggtitle("Series after Box-Cox (Dogecoin)") +
  geom_smooth(method = "loess", level = 0)
```

Comparing the variances of original Ethereum data and transformed data
```{r}
var(close_eth)
var(ethereum$close_transf_eth)
```

Comparing the variances of original Dogecoin data and transformed data
```{r}
var(close_doge)
var(dogecoin$close_transf_doge)
```

Differencing of the Ethereum data at Lag1
```{r}
par(mfrow = c(1, 1))

diff1_eth <- diff(ethereum$close_transf_eth, lag = 1)
autoplot(diff1_eth) + ggtitle("Ethereum Time Series after Differencing at lag 1 after Box-Cox")
```

Differencing of the Dogecoin data at Lag1
```{r}
par(mfrow = c(1, 1))

diff1_doge <- diff(dogecoin$close_transf_doge, lag = 1)
autoplot(diff1_doge) + ggtitle("Dogecoin Time Series after Differencing at lag 1 after Box-Cox")
```

Dickey-Fuller test to determine whether the Ethereum data is stationary
```{r}
adf.test(diff1_eth, alternative = "stationary")
```

Dickey-Fuller test to determine whether the Dogecoin data is stationary
```{r}
adf.test(diff1_doge, alternative = "stationary")
```

Fitting ARIMA model for Ethereum data
```{r}
arima_eth <- auto.arima(close_eth, stepwise = F, approximation = F, lambda = 0, ic = "aicc")
arima_eth
```
Accuracy for Arima model fitted on Ethereum dataset
```{r}
accuracy(arima_eth)
```

Fitting ARIMA model for Dogecoin data
```{r}
arima_doge <- auto.arima(close_doge, stepwise = F, approximation = F, lambda = 0, ic = "aicc")
arima_doge
```

Accuracy for Arima model fitted on Ethereum dataset
```{r}
accuracy(arima_doge)
```

Checking residuals for Ethereum and Dogecoin
```{r setup, warning=F}
k <- round(sqrt(length(open)))
# a matrix comparing p-values for the three different tests
diag.mat <- matrix(
  data = NA, nrow = 2, ncol = 3,
  dimnames = list(
    c("Ethereum", "Dogecoin"),
    c("Ljung-Box", "McLeod-Li", "Shapiro-Wilk")
  )
)

residuals_eth <- arima_eth$residuals
residuals_doge <- arima_doge$residuals

# testing Ethereum residuals
diag.mat[1, 1] <- Box.test(residuals_eth, type = "Ljung-Box", lag = k, fitdf = 0)$p.value
diag.mat[1, 2] <- Box.test(residuals_eth^2, type = "Ljung-Box", lag = k, fitdf = 0)$p.value
diag.mat[1, 3] <- shapiro.test(residuals_eth)$p.value

# testing Dogecoin residuals
diag.mat[2, 1] <- Box.test(residuals_doge, type = "Ljung-Box", lag = k, fitdf = 2)$p.value
diag.mat[2, 2] <- Box.test(residuals_doge^2, type = "Ljung-Box", lag = k, fitdf = 2)$p.value
diag.mat[2, 3] <- shapiro.test(residuals_doge)$p.value

# matrix of p-values
round(diag.mat, 3)

```


# Forecasting
```{r}
forecast_eth <- forecast(ethereum$Close, model = arima_eth, lambda = 0)
forecast_doge <- forecast(dogecoin$Close, model = arima_doge, lambda = 0)

plot(forecast_eth,main="Ethereum Coin Forecasting")
plot(forecast_doge,main="Dogecoin Forecasting")
```



