
---
title: "Time series decomposition and transformations. Practice."
date: '2022-09-21'
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, warning=F, message=F}


library(tibble)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

# tsibble: tidy temporal data frames and tools
library(tsibble)

# fable (forecast table)
library(fable)

# fabletools - provides tools for building modelling packages, with a focus on time series forecasting
library(fabletools)

# Feature Extraction and Statistics for Time Series in tsibble format
library(feasts)

# tsibbledata: used datasets for example global_economy
library(tsibbledata)

library(fpp3)
```


There are multiple libraries loaded and to make life easier here are
some associations between used function and library name:

"Regular" tydiverce

 * tidyr
   * pivot_longer - Pivot data from wide to long

Time-Series tydiverce

 * tsibble - tidy temporal data frames and tools
   * index_by - set time index
 * fable - forecast table
 * fabletools - provides tools for building modelling packages, with a focus on time series forecasting
   * autoplot - also ggplot2 and feasts depending on object
   * autolayer - also ggplot2 and feasts depending on object
   * features - Extract features from a dataset
   * model - Estimate models
 * feasts - Feature Extraction and Statistics for Time Series in tsibble format
   * autoplot - also ggplot2 and feasts depending on object
   * autolayer - also ggplot2 and feasts depending on object
   * guerrero - Guerrero's method for Box Cox lambda selection
   * STL - Multiple seasonal decomposition by Loess
   * gg_subseries
 
 
# Adjustment

```{r}
## GDP --------------------------------------------------------------------------
global_economy %>%
  filter(Country == "Australia") %>%
  autoplot(GDP)

global_economy %>%
  filter(Country == "Australia") %>%
  autoplot(GDP / Population)
```

```{r}
## Print retail adjusted by CPI --------------------------------------------------
print_retail <- aus_retail %>%
  filter(Industry == "Newspaper and book retailing") %>%
  group_by(Industry) %>%
  index_by(Year = year(Month)) %>%
  summarise(Turnover = sum(Turnover))

print_retail %>% autoplot(Turnover)
```

```{r}
aus_economy <- global_economy %>%
  filter(Code == "AUS")

print_retail <- print_retail %>%
  left_join(aus_economy, by = "Year") %>%
  mutate(Adj_turnover = Turnover / CPI * 100) %>%
  pivot_longer(c(Turnover, Adj_turnover),
    names_to = "Type", values_to = "Turnover"
  )
```

```{r}
# Plot both on same graph
print_retail %>%
  ggplot(aes(x = Year, y = Turnover, col = Type)) +
  geom_line() +
  labs(
    title = "Turnover: Australian print media industry",
    y = "$AU"
  )
```

```{r}
# Use faceting
print_retail %>%
  mutate(Type = factor(Type,
    levels = c("Turnover", "Adj_turnover")
  )) %>%
  ggplot(aes(x = Year, y = Turnover)) +
  geom_line() +
  facet_grid(Type ~ ., scales = "free_y") +
  labs(
    title = "Turnover: Australian print media industry",
    y = "$AU"
  )
```

# Math Transformation


```{r}
## Australian food retail --------------------------------------------------------

food <- aus_retail %>%
  filter(Industry == "Food retailing") %>%
  summarise(Turnover = sum(Turnover))
```

```{r}
food %>% autoplot(Turnover) +
  labs(y = "Turnover ($AUD)")
```

```{r}
food %>% autoplot(sqrt(Turnover)) +
  labs(y = "Square root turnover")
```

```{r}
food %>% autoplot(log(Turnover)) +
  labs(y = "Log turnover")
```

```{r}
food %>%
  features(Turnover, features = guerrero)
```

```{r}
food %>% autoplot(box_cox(Turnover, 0.08952696)) +
  labs(y = "Box-Cox transformed turnover")
```
# Decomposition

```{r}
## US retail employment ----------------------------------------------------------

us_retail_employment <- us_employment %>%
  filter(year(Month) >= 1990, Title == "Retail Trade") %>%
  select(-Series_ID)

us_retail_employment %>%
  autoplot(Employed) +
  labs(
    y = "Persons (thousands)",
    title = "Total employment in US retail"
  )
```

```{r}
dcmp <- us_retail_employment %>%
  model(stl = STL(Employed))
components(dcmp)
# %>% head()
```

```{r}
us_retail_employment %>%
  autoplot(Employed, color = "gray") +
  autolayer(components(dcmp), trend, color = "red") +
  labs(
    y = "Persons (thousands)",
    title = "Total employment in US retail"
  )
```

```{r}
components(dcmp) %>% autoplot()
```

```{r}
components(dcmp) %>% gg_subseries(season_year)
```

```{r}
us_retail_employment %>%
  autoplot(Employed, color = "gray") +
  autolayer(components(dcmp), season_adjust, color = "blue") +
  labs(
    y = "Persons (thousands)",
    title = "Total employment in US retail"
  )
```

```{r}
us_retail_employment %>%
  model(STL(Employed ~ season(window = 13) + trend(window = 7), robust = TRUE)) %>%
  components() %>%
  autoplot() +
  labs(title = "STL decomposition: US retail employment")
```

