---
title: "Ch5. The forecasters' toolbox. Practice."

output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, warning=F, message=F}


library(tibble)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

# tsibble: tidy temporal data frames and tools
library(tsibble)

# fable (forecast table)
library(fable)

# fabletools - provides tools for building modelling packages, with a focus on time series forecasting
library(fabletools)

# Feature Extraction and Statistics for Time Series in tsibble format
library(feasts)

# tsibbledata: used datasets for example global_economy
library(tsibbledata)

library(fpp3)
```


There are multiple libraries loaded and to make life easier here are
some associations between used function and library name:

"Regular" tydiverce
 * dplyr
   * pull=$
 * tidyr
   * pivot_longer - Pivot data from wide to long
 * generics
   * augment - Augment data with information from an object

Time-Series tydiverce

 * tsibble - tidy temporal data frames and tools
   * index_by - set time index
   * stretch_tsibble Perform stretching windows on a tsibble by row
 * fable - forecast table
   * SNAIVE, NAIVE, RW, MEAN - simple models
 * fabletools - provides tools for building modelling packages, with a focus on time series forecasting
   * autoplot - also ggplot2 and feasts depending on object
   * autolayer - also ggplot2 and feasts depending on object
   * features - Extract features from a dataset
   * model - Estimate models
   * forecast - Produce forecasts
   * features - Extract features from a dataset
   * accuracy
 * feasts - Feature Extraction and Statistics for Time Series in tsibble format
   * autoplot - also ggplot2 and feasts depending on object
   * autolayer - also ggplot2 and feasts depending on object
   * guerrero - Guerrero's method for Box Cox lambda selection
   * STL - Multiple seasonal decomposition by Loess
   * gg_subseries
   * ljung_box - Portmanteau tests
 * distributional (used by ) distributional: Vectorised Probability Distributions
   * hilo - Compute intervals
   
 


```{r}
## ---- GDP ----------------------------------------------------------------

gdppc <- global_economy %>%
  mutate(GDP_per_capita = GDP / Population) %>%
  select(Year, Country, GDP, Population, GDP_per_capita)
gdppc
```

```{r}
gdppc %>%
  filter(Country == "Sweden") %>%
  autoplot(GDP_per_capita) +
  labs(title = "GDP per capita for Sweden", y = "$US")
```

```{r}
# That is linear model we'll talk about it in next lecture
# model:
#   TSLM() (time series linear model), 
#   the response variable is GDP_per_capita
#   it is being modelled using trend() (a “special” function specifying a linear trend when it is used within TSLM())
fit <- gdppc %>%
  model(trend_model = TSLM(GDP_per_capita ~ trend()))

fc <- fit %>% forecast(h = "3 years")

fc %>%
  filter(Country == "Sweden") %>%
  autoplot(gdppc) +
  labs(title = "GDP per capita for Sweden", y = "$US")
```
# Simple Forecasting
```{r}
## ---- Bricks ------------------------------------------------------------

brick_fit <- aus_production %>%
  filter(!is.na(Bricks)) %>%
  model(
    Seasonal_naive = SNAIVE(Bricks),
    Naive = NAIVE(Bricks),
    Drift = RW(Bricks ~ drift()),
    Mean = MEAN(Bricks)
  )
```

```{r}
brick_fc <- brick_fit %>%
  forecast(h = "5 years")
```

```{r}
brick_fc %>%
  filter(.model == "Seasonal_naive") %>%
  autoplot(aus_production)
```

```{r}
z <- brick_fc %>%
  hilo(level = 95) %>%
  pull(`95%`)
z$lower

brick_fc %>%
  autoplot(aus_production, level = NULL) +
  labs(
    title = "Clay brick production for Australia",
    y = "Millions of bricks"
  ) +
  guides(colour = guide_legend(title = "Forecast"))
```

```{r}
## ---- Facebook -------------------------------------------------------------------

# Extract training data
fb_stock <- gafa_stock %>%
  filter(Symbol == "FB") %>%
  mutate(trading_day = row_number()) %>%
  update_tsibble(index = trading_day, regular = TRUE)

fb_stock %>% autoplot(Close) +
  labs(
    title = "Facebook closing stock price",
    y = "$US"
  )
```

```{r}
# Specify, estimate and forecast
fb_stock %>%
  model(
    Mean = MEAN(Close),
    Naive = NAIVE(Close),
    Drift = RW(Close ~ drift())
  ) %>%
  forecast(h = 42) %>%
  autoplot(fb_stock, level = NULL) +
  labs(
    title = "Facebook closing stock price",
    y = "$US"
  ) +
  guides(colour = guide_legend(title = "Forecast"))
```

```{r}
fit <- fb_stock %>% model(NAIVE(Close))
```

```{r}
augment(fit) %>%
  filter(trading_day > 1100) %>%
  ggplot(aes(x = trading_day)) +
  geom_line(aes(y = Close, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted"))
```

```{r}
augment(fit) %>%
  autoplot(.resid) +
  labs(
    y = "$US",
    title = "Residuals from naïve method"
  )
```

```{r}
augment(fit) %>%
  ggplot(aes(x = .resid)) +
  geom_histogram(bins = 150) +
  labs(title = "Histogram of residuals")
```

```{r}
augment(fit) %>%
  ACF(.resid) %>%
  autoplot() +
  labs(title = "ACF of residuals")
```

```{r}
gg_tsresiduals(fit)
```

```{r}
augment(fit) %>%
  features(.resid, ljung_box, lag = 10, dof = 0)

fc <- fb_stock %>%
  model(
    Mean = MEAN(Close),
    Naive = NAIVE(Close),
    Drift = RW(Close ~ drift())
  ) %>%
  forecast(h = 42)
```

```{r}
## EGG PRICES

eggs <- prices %>%
  filter(!is.na(eggs)) %>%
  select(eggs)
```

```{r}
eggs %>%
  autoplot(log(eggs)) +
  labs(
    title = "Annual egg prices",
    y = "$US (adjusted for inflation)"
  )
```

```{r}
fit <- eggs %>%
  model(rwdrift = RW(log(eggs) ~ drift()))
fit
```

```{r}
fc <- fit %>%
  forecast(h = 50)
fc
```

```{r}
fc %>% autoplot(eggs) +
  labs(
    title = "Annual egg prices",
    y = "US$ (adjusted for inflation)"
  )
```

```{r}
fc %>%
  autoplot(eggs, level = 80, point_forecast = lst(mean, median)) +
  labs(
    title = "Annual egg prices",
    y = "US$ (adjusted for inflation)"
  )
```

```{r}

## US RETAIL EMPLOYMENT

us_retail_employment <- us_employment %>%
  filter(year(Month) >= 1990, Title == "Retail Trade") %>%
  select(-Series_ID)
```

```{r}
us_retail_employment %>%
  autoplot(Employed)
```

```{r}
dcmp <- us_retail_employment %>%
  model(STL(Employed)) %>%
  components()
```

```{r}
autoplot(dcmp)
```

```{r}
dcmp <- dcmp %>% select(-.model)
dcmp %>% autoplot(season_adjust)
```

```{r}
dcmp %>%
  model(NAIVE(season_adjust)) %>%
  forecast() %>%
  autoplot(dcmp) +
  labs(title = "Naive forecasts of seasonally adjusted data")
```

```{r}
dcmp %>% autoplot(season_year)
```

```{r}
dcmp %>%
  model(SNAIVE(season_year)) %>%
  forecast() %>%
  autoplot(dcmp) +
  labs(title = "Seasonal naive forecasts of seasonal component")
```

```{r}
us_retail_employment %>%
  model(stlf = decomposition_model(
    STL(Employed),
    NAIVE(season_adjust),
    SNAIVE(season_year)
  )) %>%
  forecast() %>%
  autoplot(us_retail_employment)
```

# Accuracy
```{r}
## BEER PRODUCTION

recent_production <- aus_production %>%
  filter(year(Quarter) >= 1992)
```

```{r}
recent_production %>% autoplot(Beer)
```

```{r}
train <- recent_production %>%
  filter(year(Quarter) <= 2007)
```

```{r}
beer_fit <- train %>%
  model(
    Mean = MEAN(Beer),
    Naive = NAIVE(Beer),
    Seasonal_naive = SNAIVE(Beer),
    Drift = RW(Beer ~ drift())
  )
```

```{r}
beer_fc <- beer_fit %>%
  forecast(h = 10)
```

```{r}
accuracy(beer_fc, recent_production)
```

```{r}
accuracy(beer_fit)
```

# CROSS-VALIDATION

```{r}
## CROSS-VALIDATION: FACEBOOK

fb_stock <- gafa_stock %>%
  filter(Symbol == "FB") %>%
  mutate(trading_day = row_number()) %>%
  update_tsibble(index = trading_day, regular = TRUE)
```

```{r}
fb_stock %>%
  autoplot(Close)
```

```{r}
fb_stretch <- fb_stock %>%
  stretch_tsibble(.init = 3, .step = 1) %>%
  filter(.id != max(.id))
```

```{r}
fit_cv <- fb_stretch %>%
  model(RW(Close ~ drift()))
```

```{r}
fc_cv <- fit_cv %>%
  forecast(h = 1)
```

```{r}
fc_cv %>% accuracy(fb_stock)
```

```{r}
fb_stock %>%
  model(RW(Close ~ drift())) %>%
  accuracy()
```
