---
title: "Ch7. Regression model. Practice."

output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, warning=F, message=F}


library(tibble)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

# tsibble: tidy temporal data frames and tools
library(tsibble)

# fable (forecast table)
library(fable)

# fabletools - provides tools for building modelling packages, with a focus on time series forecasting
library(fabletools)

# Feature Extraction and Statistics for Time Series in tsibble format
library(feasts)

# tsibbledata: used datasets for example global_economy
library(tsibbledata)

library(fpp3)
```


There are multiple libraries loaded and to make life easier here are
some associations between used function and library name:

"Regular" tydiverce
 * dplyr
   * pull=$
 * tidyr
   * pivot_longer - Pivot data from wide to long
 * generics
   * augment - Augment data with information from an object

Time-Series tydiverce

 * tsibble - tidy temporal data frames and tools
   * index_by - set time index
   * stretch_tsibble Perform stretching windows on a tsibble by row
 * fable - forecast table
   * SNAIVE, NAIVE, RW, MEAN - simple models
 * fabletools - provides tools for building modelling packages, with a focus on time series forecasting
   * autoplot - also ggplot2 and feasts depending on object
   * autolayer - also ggplot2 and feasts depending on object
   * features - Extract features from a dataset
   * model - Estimate models
   * forecast - Produce forecasts
   * features - Extract features from a dataset
   * accuracy
 * feasts - Feature Extraction and Statistics for Time Series in tsibble format
   * autoplot - also ggplot2 and feasts depending on object
   * autolayer - also ggplot2 and feasts depending on object
   * guerrero - Guerrero's method for Box Cox lambda selection
   * STL - Multiple seasonal decomposition by Loess
   * gg_subseries
   * ljung_box - Portmanteau tests
 * distributional (used by ) distributional: Vectorised Probability Distributions
   * hilo - Compute intervals
   
```{r}
# US consumption quarterly changes

us_change %>%
  pivot_longer(-Quarter, names_to = "Measure", values_to = "Change") %>%
  ggplot(aes(x = Quarter, y = Change, colour = Measure)) +
  geom_line() +
  facet_grid(vars(Measure), scales = "free_y") +
  labs(y = "") +
  guides(colour = "none")
```

```{r}
us_change %>%
  GGally::ggpairs(columns = 2:6)
```

```{r}
fit_consMR <- us_change %>%
  model(lm = TSLM(Consumption ~ Income + Production + Unemployment + Savings))
report(fit_consMR)
```

```{r}
augment(fit_consMR) %>%
  ggplot(aes(x = Quarter)) +
  geom_line(aes(y = Consumption, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted")) +
  labs(
    y = NULL,
    title = "Percent change in US consumption expenditure"
  ) +
  scale_colour_manual(values = c(Data = "black", Fitted = "#D55E00")) +
  guides(colour = guide_legend(title = NULL))
```

```{r}
augment(fit_consMR) %>%
  ggplot(aes(y = .fitted, x = Consumption)) +
  geom_point() +
  labs(
    y = "Fitted (predicted values)",
    x = "Data (actual values)",
    title = "Percentage change in US consumption expenditure"
  ) +
  geom_abline(intercept = 0, slope = 1)

fit_consMR %>% gg_tsresiduals()
```

```{r}
# Australian beer production

recent_production <- aus_production %>% filter(year(Quarter) >= 1992)
recent_production %>%
  autoplot(Beer) +
  labs(y = "Megalitres", title = "Australian quarterly beer production")
```

```{r}
fit_beer <- recent_production %>%
  model(TSLM(Beer ~ trend() + season()))
report(fit_beer)
```

```{r}
augment(fit_beer) %>%
  ggplot(aes(x = Quarter)) +
  geom_line(aes(y = Beer, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted")) +
  labs(y = "Megalitres", title = "Australian quarterly beer production") +
  scale_colour_manual(values = c(Data = "black", Fitted = "#D55E00"))
```

```{r}
augment(fit_beer) %>%
  ggplot(aes(x = Beer, y = .fitted, colour = factor(quarter(Quarter)))) +
  geom_point() +
  labs(
    y = "Fitted", x = "Actual values",
    title = "Quarterly beer production"
  ) +
  scale_colour_brewer(palette = "Dark2", name = "Quarter") +
  geom_abline(intercept = 0, slope = 1)
```

```{r}
fit_beer %>% gg_tsresiduals()
```

```{r}
fit_beer %>%
  forecast() %>%
  autoplot(recent_production)
```

```{r}
fourier_beer <- recent_production %>%
  model(TSLM(Beer ~ trend() + fourier(K = 2)))
report(fourier_beer)
```

```{r}
recent_production %>%
  model(
    f1 = TSLM(Beer ~ trend() + fourier(K = 1)),
    f2 = TSLM(Beer ~ trend() + fourier(K = 2)),
    season = TSLM(Beer ~ trend() + season())
  ) %>%
  glance()

```

```{r}
# Fourier terms for cafe data

aus_cafe <- aus_retail %>%
  filter(
    Industry == "Cafes, restaurants and takeaway food services",
    year(Month) %in% 2004:2018
  ) %>%
  summarise(Turnover = sum(Turnover))
aus_cafe %>%
  autoplot(Turnover)
aus_cafe %>%
  autoplot(log(Turnover))
```

```{r}
fit <- aus_cafe %>%
  model(
    K1 = TSLM(log(Turnover) ~ trend() + fourier(K = 1)),
    K2 = TSLM(log(Turnover) ~ trend() + fourier(K = 2)),
    K3 = TSLM(log(Turnover) ~ trend() + fourier(K = 3)),
    K4 = TSLM(log(Turnover) ~ trend() + fourier(K = 4)),
    K5 = TSLM(log(Turnover) ~ trend() + fourier(K = 5)),
    K6 = TSLM(log(Turnover) ~ trend() + fourier(K = 6))
  )
```

```{r}
augment(fit) %>%
  filter(.model %in% c("K1", "K2", "K3")) %>%
  ggplot(aes(x = Month, y = Turnover)) +
  geom_line() +
  geom_line(aes(y = .fitted, col = .model)) +
  facet_grid(.model ~ .)
```

```{r}
glance(fit) %>%
  select(.model, sigma2, log_lik, AIC, AICc, BIC)

```

```{r}
# Boston Marathon

marathon <- boston_marathon %>%
  filter(Event == "Men's open division") %>%
  select(-Event) %>%
  mutate(Minutes = as.numeric(Time) / 60)
```

```{r}
marathon %>%
  autoplot(Minutes) +
  labs(y = "Winning times in minutes")
```

```{r}
fit_trends <- marathon %>%
  model(
    # Linear trend
    linear = TSLM(Minutes ~ trend()),
    # Exponential trend
    exponential = TSLM(log(Minutes) ~ trend()),
    # Piecewise linear trend
    piecewise = TSLM(log(Minutes) ~ trend(knots = c(1940, 1980)))
  )
```

```{r}
fit_trends

```

```{r}
fit_trends %>%
  select(piecewise) %>%
  report()
```

```{r}
fc_trends <- fit_trends %>%
  forecast(h = 10)
```

```{r}
marathon %>%
  autoplot(Minutes) +
  geom_line(
    data = fitted(fit_trends),
    aes(y = .fitted, colour = .model)
  ) +
  autolayer(fc_trends, alpha = 0.5, level = 95) +
  labs(
    y = "Minutes",
    title = "Boston marathon winning times"
  )
```

```{r}
fit_trends %>%
  select(piecewise) %>%
  gg_tsresiduals()
```

```{r}
glance(fit_trends) %>%
  select(.model, r_squared, adj_r_squared, AICc, CV)

```

```{r}
# US consumption quarterly changes

fit_all <- us_change %>%
  model(
    TSLM(Consumption ~ Income + Production + Unemployment + Savings),
    TSLM(Consumption ~ Production + Unemployment + Savings),
    TSLM(Consumption ~ Income + Unemployment + Savings),
    TSLM(Consumption ~ Income + Production + Savings),
    TSLM(Consumption ~ Income + Production + Unemployment),
    TSLM(Consumption ~ Income + Production),
    TSLM(Consumption ~ Income + Unemployment),
    TSLM(Consumption ~ Income + Savings),
    TSLM(Consumption ~ Production + Unemployment),
    TSLM(Consumption ~ Production + Savings),
    TSLM(Consumption ~ Unemployment + Savings),
    TSLM(Consumption ~ Income),
    TSLM(Consumption ~ Production),
    TSLM(Consumption ~ Unemployment),
    TSLM(Consumption ~ Savings),
    TSLM(Consumption ~ 1),
  )
```

```{r}
glance(fit_all) %>%
  select(.model, AICc, CV) %>%
  arrange(AICc)
```

```{r}

us_change %>%
  model(
    TSLM(Consumption ~ Income * Savings + Production + Unemployment),
  ) %>%
  report()
```

```{r}
fit_all %>%
  glance() %>%
  select(.model, adj_r_squared, AICc, BIC, CV) %>%
  arrange(CV)
```

```{r}
fit_consBest <- us_change %>%
  model(
    TSLM(Consumption ~ Income + Production + Unemployment + Savings),
  )
```

```{r}
fit_consBest %>% report()
```

```{r}
future_scenarios <- scenarios(
  Increase = new_data(us_change, 4) %>%
    mutate(Income = 1, Savings = 0.5, Unemployment = 0, Production = 0),
  Decrease = new_data(us_change, 4) %>%
    mutate(Income = -1, Savings = -0.5, Unemployment = 0, Production = 0),
  names_to = "Scenario"
)
```

```{r}
fc <- forecast(fit_consBest, new_data = future_scenarios)
```

```{r}
us_change %>% autoplot(Consumption) +
  labs(y = "% change in US consumption") +
  autolayer(fc) +
  labs(title = "US consumption", y = "% change")
```