---
title: "Time Series Modeling. Practice with_tsible"
author: "Author"
date: "2022-11-06"
output: html_document
---


```{r setup, warning=F, message=F}
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)

# tsibble: tidy temporal data frames and tools
library(tsibble)

# fable (forecast table)
library(fable)

# fabletools - provides tools for building modelling packages, with a focus on time series forecasting
library(fabletools)

# Feature Extraction and Statistics for Time Series in tsibble format
library(feasts)

# tsibbledata: used datasets for example global_economy
library(tsibbledata)
library(cowplot)
```

# Read from file

```{r}
df <- readr::read_csv("antidiabetic_drug_sales.csv",show_col_types = FALSE)
head(df)
```

# Convert to tsible

```{r}
df %>% 
  mutate(Month=yearmonth(Month)) %>% 
  tsibble(index=Month) -> 
  df
head(df)
```

```
other helpful function for time to index conversion
Frequency	Function

Annual  	start:end
Quarterly	yearquarter()
Monthly   yearmonth()
Weekly    yearweek()
Daily	    as_date(), ymd()
Sub-daily	as_datetime()
```

# Plotting


```{r}
# Split to train and test (whole = train + test)
df_whole <- df
df <- df %>% filter(Month < yearmonth("2005-01"))
```

```{r}
# plot
autoplot(df,Cost)
```





Seasonal plot
```{r, echo=TRUE, dependson="a10", fig.height=3.3}
df %>% gg_season(Cost, labels = "both") +
  labs(y = "$ million", title = "Seasonal plot: antidiabetic drug sales")
```


Seasonal subseries plots

```{r}
df %>%
  gg_subseries(Cost) +
  labs(y = "$ million", title = "Subseries plot: antidiabetic drug sales")
```
```{r}
# ACF(df, Cost) %>% autoplot()
# PACF(df, Cost) %>% autoplot()
gg_tsdisplay(df, Cost, plot_type='partial')
```

# Trend-Cycle / Season decomposition

```
classical_decomposition(formula, type = c("additive", "multiplicative"), ...)
STL(formula, iterations = 2, ...)
    ... - stata:stl options
```

```{r}
df %>% model(classical_decomposition(Cost, type='m')) %>% components() %>% autoplot()
```

Transformation?

```{r}
df %>% model(classical_decomposition(log(Cost), type='m')) %>% components() %>% autoplot()
```

# Transformation

```{r}
# Auto Box-Cox
# log if lambda is zero
df %>% features(Cost, features = guerrero)
```

```{r}
plot_grid(
  autoplot(df,Cost),
  autoplot(df,box_cox(Cost,0.272)),
  autoplot(df,log(Cost)),
  ncol=3)
```


# Simple models


```{r}
# fit
fit <- df %>%
  model(
    Mean = MEAN(Cost),
    Naive = NAIVE(Cost),
    Seasonal_Naive = SNAIVE(Cost),
    Drift = RW(Cost ~ drift())
  )
accuracy(fit)
```

```{r}
# forecast next year (in training set)
fc <- fit %>% forecast(h = 36)
# plot forecasts and actual data
fc %>% autoplot(df_whole,level = NULL)
```

```{r}
# Calculate accuracy
accuracy(fc,df_whole)
```

Confidence intervals?
```{r}
# forecast next year (in training set)
fc <- fit %>% forecast(h = 36)
#plot forecasts and actual data
fc %>% autoplot(df_whole,level = 80)
fc %>% autoplot(df_whole,level = 80) + facet_wrap(~.model)
```

```{r}
gg_tsresiduals(fit %>% select(Drift))
gg_tsresiduals(fit %>% select(Seasonal_Naive))
```
```{r}
# lag=10 for non-seasonal data, 
# lag=2m for seasonal data (where m is seasonal period).

augment(fit) %>%  features(.resid, ljung_box, lag=24, dof=1)

```

May be need transform
```{r}
# fit
fit <- df %>%
  model(
    Mean = MEAN(log(Cost)),
    Naive = NAIVE(log(Cost)),
    Seasonal_Naive = SNAIVE(log(Cost)),
    Drift = RW(log(Cost) ~ drift())
  )
accuracy(fit)
```

```{r}
# forecast next year (in training set)
fc <- fit %>% forecast(h = 36)
#plot forecasts and actual data
fc %>% autoplot(df_whole,level = 80)
fc %>% autoplot(df_whole,level = 80, point_forecast = lst(mean, median))
fc %>% autoplot(df_whole,level = 80) + facet_wrap(~.model)
```

```{r}
fc %>% hilo(Cost,level = 95) %>%
  unpack_hilo(`95%`)
```


```{r}
gg_tsresiduals(fit %>% select(Drift))
gg_tsresiduals(fit %>% select(Seasonal_Naive))
```

```{r}
# lag=10 for non-seasonal data, 
# lag=2m for seasonal data (where m is seasonal period).

augment(fit) %>%  features(.innov, ljung_box, lag=24, dof=1)

```

# EST
```{r}
df %>% model(STL(log(Cost))) %>% components() %>% autoplot()
```

```{r}
fit <- df %>%
  model(
    ets_auto = ETS(log(Cost)),
    ets = ETS(log(Cost) ~ error("A") + trend("A") + season("A"))
  )
accuracy(fit)
report(fit)
report(fit[1])

fit <- fit %>% select(ets)

fc <- fit %>% forecast(h = "3 years")
```

```{r}
fc %>% autoplot(df_whole,level = 90)
```

```{r}
accuracy(fc,df_whole)
```
```{r}
gg_tsresiduals(fit)
```

```{r}
# lag=10 for non-seasonal data, 
# lag=2m for seasonal data (where m is seasonal period).

augment(fit) %>%  features(.innov, ljung_box, lag=24, dof=17)

```
# ARIMA
```{r}
gg_tsdisplay(df, Cost, plot_type='partial')
```

```{r}
# seasonal?
gg_tsdisplay(df, difference(Cost), plot_type='partial')
```
```{r}
df %>%
features(Cost, unitroot_kpss)
```

```{r}
df %>% features(Cost, unitroot_nsdiffs)
df %>% features(log(Cost), unitroot_nsdiffs)
```
```{r}
df %>% features( difference(log(Cost), lag=12), unitroot_ndiffs)
```
AR(p)
* ACF dies out in an exponential or damped sine-wave manner
* PACF has all zero spikes beyond the 𝑝th spike
So we have an AR(p) model when
* the ACF is exponentially decaying or sinusoidal
* there is a significant spike at lag p in PACF, but none beyond p
MA(q)
* PACF dies out in an exponential or damped sine-wave manner
* ACF has all zero spikes beyond the 𝑞th spike
So we have an MA(q) model when
* the PACF is exponentially decaying or sinusoidal
* there is a significant spike at lag q in ACF, but none beyond q

```{r}
gg_tsdisplay(df, difference(log(Cost), lag=12), plot_type='partial')
```
p=3 q=3 and seasonal ar(1)

```{r}
fit <- df %>%
  model(
    arima_auto = ARIMA(log(Cost)),
    arima = ARIMA(log(Cost)~0+pdq(3,0,3)+PDQ(1,1,0))
  )
accuracy(fit)
report(fit[1])
report(fit[2])

fc <- fit %>% forecast(h = "3 year")
```

```{r}
fc %>% autoplot(df_whole,level = 80)
```

```{r}
accuracy(fc,df_whole)
```

```{r}
gg_tsresiduals(fit %>% select(arima_auto))
```

```{r}
# lag=10 for non-seasonal data, 
# lag=2m for seasonal data (where m is seasonal period).

augment(fit %>% select(arima_auto)) %>%  features(.innov, ljung_box, lag=24, dof=7)

```