---
title: "Survival Analysis - Monte Carlo Simulation Study"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)

library(ggplot2)
library(cowplot)
#library(plotly)

library(survival)
library(survminer)
```

# Survival Analysis - Monte Carlo Simulation Study

The purpose of this study is to see the effect of earlier subject dismissal on accuracy of survival at time of study completion depending on simple calculation or survival probability chaining.


# Distribution Selection

```{r }
# BrainCancer data set
library(ISLR2)

ggplot(BrainCancer,aes(x=time)) + facet_wrap(~status) + geom_histogram()
```

```{r }
set.seed(4202022)
# Poisson
df <- rbind(
  data.frame(lambda=10,t=rpois(5000,10)),
  data.frame(lambda=20,t=rpois(5000,20)),
  data.frame(lambda=40,t=rpois(5000,40))
)
ggplot(df, aes(x=t)) + facet_wrap(~lambda,ncol=1) + geom_histogram(binwidth=1)
```


```{r }
set.seed(4202022)

df <- rbind(
  data.frame(df=10,t=rchisq(5000,10)),
  data.frame(df=20,t=rchisq(5000,20)),
  data.frame(df=40,t=rchisq(5000,40))
)
ggplot(df, aes(x=t)) + facet_wrap(~df,ncol=1) + geom_histogram(binwidth=1)
```
```{r }
set.seed(4202022)

df <- NULL
for(shape in c(1,2,4)){
  for(scale in c(0.5, 1, 2)){
    df <- rbind(
      df,
      data.frame(shape=shape,scale=scale,t=rgamma(5000,shape=shape,scale=scale))
      )
  }
}
ggplot(df, aes(x=t)) + facet_grid(as.factor(shape)~as.factor(scale)) + geom_histogram()
```

```{r }
set.seed(4202022)

df <- NULL
for(shape in c(3, 4, 5)){
  for(scale in c(15, 10, 4)){
    df <- rbind(
      df,
      data.frame(shape=shape,scale=scale,t=rgamma(5000,shape=shape,scale=scale))
      )
  }
}
ggplot(df, aes(x=t)) + facet_grid(as.factor(shape)~as.factor(scale)) + geom_histogram()
```
Lets say whole study is 60 months

```{r }
# probability of what?
pgamma(60, shape=4,scale=10)
```
Lets event been gamma distributed with shape=4,scale=15

Lets study drop out be calculated as:

1. Select n_d drop outs
2. Generate drop out time uniformly from start of study to the end
3. If drop out time smaller than event, go with it otherwise skip it

```{r }
set.seed(4202022)

# number of subjects
n <- 1000
# number of possible droup out
n_d <- as.integer(0.7*n)
# end of study
t_end <- 60

shape<-4
scale<-10

df <- data.frame(t_event=rgamma(n,shape=shape,scale=scale), drop_outs_time = 2*t_end, t=NA, status=1)

drop_outs_id <- sample(n,n_d)
df$drop_outs_time[drop_outs_id] <- runif(n_d,0,t_end)

# apply censoring
df$drop_outs_time[df$drop_outs_time > t_end] <- t_end

drop_outs_id <- df$drop_outs_time < df$t_event
df$status[drop_outs_id] <- 0

df$t <- df$t_event
df$t[drop_outs_id] <- df$drop_outs_time[drop_outs_id]

cat("drop outs (%):", 100*sum(df$status==0)/n)
head(df)
```


```{r}
plot_grid(
  ggplot(df %>% filter(t<60 & status==0),aes(x=t)) + geom_histogram(breaks=c(seq(0,150,5))),
  ggplot(df %>% filter(t<60 & status==1),aes(x=t)) + geom_histogram(breaks=c(seq(0,150,5))),
  ggplot(df, aes(x=t_event)) + geom_histogram(breaks=c(seq(0,150,5))),
  ncol=1
)
```
```{r}
cat("Try survival to end of study\n")
1 - pgamma(t_end, shape=shape,scale=scale)


cat("Simple Binomial calculation\n")
event_occurs <- filter(df, status==1 & t <= 60) %>% nrow()
dropouts <- filter(df, status==0 & t < 60) %>% nrow()
no_event <- filter(df, t >= 60) %>% nrow()
cat("event_occurs",event_occurs,"dropouts",dropouts,"no_event",no_event,"\n")

binom.test(c(no_event,event_occurs))


cat("Survival\n")
fit_surv <- survfit(Surv(t, status) ~ 1, data=df)
summary(fit_surv, times = 60)

ggsurvplot(fit = fit_surv)
```

Method    | Value
----------|------
Try       | 0.15
Simple    | 0.08
Surv. An. | 0.17

Can we generalize?

```{r}
cat("Simple Binomial calculation without censoring\n")
event_occurs <- filter(df, t_event <= 60) %>% nrow()
no_event <- filter(df,t_event >= 60) %>% nrow()

binom.test(c(no_event,event_occurs))
```


# Lets get some stats

```{r}
set.seed(4202022)

gen_sample <- function(n,n_d,t_end=60, shape=4, scale=10, run_id=0) {
  df <- data.frame(run_id=run_id, t_event=rgamma(n,shape=shape,scale=scale), drop_outs_time = 2*t_end, t=NA, status=1)
  
  drop_outs_id <- sample(n,n_d)
  df$drop_outs_time[drop_outs_id] <- runif(n_d,0,t_end)
  
  # apply censoring
  df$drop_outs_time[df$drop_outs_time > t_end] <- t_end
  
  drop_outs_id <- df$drop_outs_time < df$t_event
  df$status[drop_outs_id] <- 0
  
  df$t <- df$t_event
  df$t[drop_outs_id] <- df$drop_outs_time[drop_outs_id]

  return(df)
}

gen_all_samples <- function(n_sim, n,n_d,t_end=60, shape=4, scale=10) {
  df <- NULL
  for(run_id in 1:n_sim){
    df <- rbind(
      df,
      gen_sample(n,n_d,t_end=t_end, shape=shape, scale=scale, run_id=run_id)
    )
  }
  return(df)
}

# number of simulations
n_sim <- 5
# number of subjects
n <- 10
# number of possible droup out
n_d <- as.integer(0.7*n)
# end of study
t_end <- 60
# event distribution
shape <-4
scale <- 10

df <- gen_all_samples(n_sim, n,n_d,t_end=60, shape=4, scale=10)
#head(df)
df
```
```{r}
# number of simulations
n_sim <- 1000
# number of subjects
n <- 100
# number of possible droup out
n_d <- as.integer(0.7*n)
# end of study
t_end <- 60
# event distribution
shape <-4
scale <- 10

df <- gen_all_samples(n_sim, n,n_d,t_end=60, shape=4, scale=10)
#head(df)
df
```

```{r}
dropout_calc <- function(m_df) {
  dropout_perc <- 100*sum(m_df$status==0 & m_df$t < t_end)/nrow(m_df)
  
  # Try survival to end of study
  theory <- 1 - pgamma(t_end, shape=shape,scale=scale)
  
  
  # Simple Binomial calculation
  event_occurs <- filter(m_df, status==1 & t <= t_end) %>% nrow()
  no_event <- filter(m_df, t >= t_end) %>% nrow()
  
  simple <- binom.test(c(no_event,event_occurs))$estimate
  
  
  # "Survival\n")
  
  fit_surv <- survfit(Surv(t, status) ~ 1, data=m_df)
  
  if(fit_surv$time[length(fit_surv$time)] < t_end) {
    # all events occur till the end of study
    survan <- 0
  }else{
    survan <- summary(fit_surv, times = 60)$surv
  }
  
  
  data.frame(
    dropout_perc=dropout_perc,
    theory=theory,
    simple=simple,
    survan=survan
    )
}
surv_dist <- df %>% group_by(run_id) %>% do(dropout_calc(.))
head(surv_dist)
```

```{r}
surv_dist %>% tidyr::gather(key = "key", value = "value",-run_id, -dropout_perc,-theory) %>%
  ggplot(aes(x=value))+facet_wrap(~key,ncol=1)+geom_histogram()+geom_vline(xintercept=surv_dist$theory[1])
```


```{r}
# number of simulations
n_sim <- 1000
# number of subjects
n <- 200
# number of possible droup out
n_d <- as.integer(0.7*n)
# end of study
t_end <- 60
# event distribution
shape <-4
scale <- 10

df <- gen_all_samples(n_sim, n,n_d,t_end=60, shape=4, scale=10)

surv_dist <- df %>% group_by(run_id) %>% do(dropout_calc(.))

surv_dist %>% tidyr::gather(key = "key", value = "value",-run_id, -dropout_perc,-theory) %>%
  ggplot(aes(x=value))+facet_wrap(~key,ncol=1)+geom_histogram()+geom_vline(xintercept=surv_dist$theory[1])
```